\documentclass[11pt]{article}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{epstopdf}
\usepackage{amsfonts}
\usepackage{natbib}
\usepackage{subfigure}
\usepackage{pdfsync}
\usepackage{xspace}
%\usepackage[pdftex, colorlinks=true,urlcolor=blue]{hyperref}
%\usepackage{wallpaper}

%%%% HERE IS SOME WATERMARK STUFF
%\addtolength{\wpXoffset}{-.2in}
%\CenterWallPaper{1.1}{/Users/eriq/Documents/work/nonprj/WaterMarks/StrictDraftWatermark.eps}


\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}


%% some handy things for making bold math
\def\bm#1{\mathpalette\bmstyle{#1}}
\def\bmstyle#1#2{\mbox{\boldmath$#1#2$}}
\newcommand{\thh}{^\mathrm{th}}


%% Some pretty etc.'s, etc...
\newcommand{\cf}{{\em cf.}\xspace }
\newcommand{\eg}{{\em e.g.},\xspace }
\newcommand{\ie}{{\em i.e.},\xspace }
\newcommand{\etal}{{\em et al.}\ }
\newcommand{\etc}{{\em etc.}\@\xspace}



%% the page dimensions from TeXShop's default---very nice
\textwidth = 6.5 in
\textheight = 9 in
\oddsidemargin = 0.0 in
\evensidemargin = 0.0 in
\topmargin = 0.0 in
\headheight = 0.0 in
\headsep = 0.0 in
\parskip = 0.2in
\parindent = 0.0in


\title{Explanation of a few modifications of\\
{\em structure} to arrive \\
 at {\em flockture}}
\author{Eric C. Anderson\thanks{
    Fisheries Ecology Division, 
    Southwest Fisheries Science Center, 
    110 Shaffer Road,
    Santa Cruz, CA 95060}
}
\begin{document}

\maketitle

\begin{abstract}
Short abstract.
\end{abstract}




\section{Adjustments to allele frequencies}

There are two main differences between {\em structure} and {\em flock} in this area.
{\em structure} samples allele frequencies from a  Dirichlet distribution conditional on a prior that has non-zero weight on every allele and 
on all the gene copies currently allocated to a particular population.  In the {\em flock} approach, by contrast, 
allele frequencies may end up being computed differently for each individual because of the leave-one-out requirement
that comes out of doing Gibbs sampling while marginalizing over the allele frequencies, and also because they do not 
choose a Dirichlet prior on allele frequencies, but rather, deal with allele frequencies of zero on the fly
in a Paetkau-like manner.

Pierre Duchesne answered my question about how the dealt with zero-frequency alleles this way:

{\sl 
Here is how FLOCK deals with zero-frequency in references:
\begin{enumerate}
\item AFLP ~~ (more generally binary data) The frequency used is $\frac{1}{n+1}$ where $n$ stands for  the size (number of genotypes) of the reference in its current state (iteration).
\item MSAT ~~ The frequency used is $\frac{1}{2n+1}$ where $n$ stands for  the size of the reference in its current state (iteration).
\end{enumerate}
To be more precise, $n$ is corrected to $n-1$ when the reference considered is that of the currently processed
genotype if the latter was part of this reference in the previous partition.
This is a consequence of the application of the leave-one-out procedure 
(all frequencies are also corrected accordingly).
 
Other commonly used rules for the zero-frequency correction would sometimes produce slightly
different cluster compositions involving only a few `gray' genotypes.
However the impact on the number and length of the plateaus would be practically null
and so the decisions from the application of the stopping and the estimation (K)
rules would be the same.
}

OK.  That seems pretty straightforward.  Here I write down the strategy for implementing this.  My main concern is going
to be to avoid lots of floating point divides during the leave-one-out part.  It will be better if we can precompute the reciprocals
so that most of those will become floating point multiplies. Some things are worth noticing:
\begin{itemize}
\item {\em structure} allows partially missing data at a locus, but I am going to implement this assuming that if one gene copy is missing at a locus
then both of them must be, since that appears to be how it is thought of in {\em flock}.
\item We only need to deal with zero-allele frequencies when we are computing the genotype probability for an individual that actually carries
that missing allele.
\end{itemize}

Index populations by $i = 1,\ldots, M$, loci by $\ell = 1,\ldots,L$, and alleles by $k = 1,\ldots, K_\ell$. Since we will only ever be focusing on
a single individual at a time, we don't need to index individuals.  Let $x_{i\ell k}$ denote the number of copies of allele $k$ at locus $j$ counted in population $i$.  With no modifications, then, we would precompute the allele frequencies in the standard fashion as:
\[
p_{i\ell k} = \frac{x_{i\ell k}}{n_{i\ell}}
\]
where $n_{i\ell}$ is the number of observed {\em gene copies} (not individuals) at locus $\ell$ in population $i$.  We will also denote the allelic type
of the two gene copies in the focal individual at locus $\ell$ by $y$ and $y'$, respectively.

Now, the adjustments to allele frequencies fall into a variety of categories as follows:
\begin{description}
	\item[Focal individual not currently allocated to $i$:] Let's see. 
	\begin{description}
		\item [Neither $x_{i\ell y}$ nor $x_{i\ell y'}$ are zero:] No adjustments necessary.  Use $p_{i\ell y}$ and $p_{i\ell y'}$ as before.
		\item [Exactly one of $x_{i\ell y}$ or $x_{i\ell y'}$ is zero:] Assume wlog that that $x_{i\ell y} = 0$.  Then, we will use:
		\begin{eqnarray*}
			p^*_{i\ell y} & = & \frac{1}{n_{i\ell} + 1} \\
			p^*_{i\ell y'} & = & \frac{x_{i\ell y'}}{n_{i\ell} + 1} = p_{i\ell y'} S_i^{(1)}
		\end{eqnarray*}
		where $S_i^{(1)} = \frac{n_i}{n_i + 1}$ and can be precomputed at the beginning of each sweep when allele frequencies are computed.
		\item [Both of $x_{i\ell y}$ or $x_{i\ell y'}$ are zero:] ~
		\begin{description}
			\item [$y$ and $y'$ are the same allele (\ie the focal indiv is homozygous):] In this case
			\[
				p^*_{i\ell y} = p^*_{i\ell y'} = \frac{1}{n_i + 1}
			\]
			and no other adjustments are necessary or relevant.  Note that $\frac{1}{n_i + 1}$ can be precomputed.
			\item [$y$ and $y'$ are different alleles:] In this case, each allele gets a count of one added to it
			so we have to add 2 instead of 1 to the normalization.
			\[
				p^*_{i\ell y} = p^*_{i\ell y'} = \frac{1}{n_i + 2}
			\]
		\end{description}
	\end{description}
	\item [Focal individual currently allocated to $i$:] In this case we have to take care of the leave-one-out aspect of things too.
	and we have to be aware that leaving an individual's genotype out might create a zero allele frequency where non had existed before.  
	Operationally, I think the first thing to do is compute what we will call $x^*$---the frequency after leaving out the focal
	individual's genotypes. 
	\begin{description}
		\item [The focal individual is homozygous:] In this case we have:
		\[
			x^*_{i\ell y} = x^*_{i\ell y'} = x_{i\ell y} - 2
		\]
		\begin{description}
			\item [If $x^*_{i\ell y} > 0$:] In this case these is no need for a zero adjustment and we just use
			\[
				p^*_{i\ell y} = p^*_{i\ell y'} = \frac{x^*_{i\ell y}}{n_i - 2} = x^*_{i\ell y} R_i^{(2)}
			\]
			where $R_i^{(2)} = \frac{1}{n_i - 2}$ is the reciprocal of $n_i - 2$, which can be precomputed.
			\item [If $x^*_{i\ell y} \leq 0$:] In this case we must make a zero adjustment by adding one copy of the
			missing allele.  So, we've removed two gene copies but then we have, in effect, put one back.  So:   
			\[
				p^*_{i\ell y} = p^*_{i\ell y'} = \frac{1}{n_i - 1} = x^*_{i\ell y} R_i^{(1)}
			\]
			where $R_i^{(1)} = \frac{1}{n_i - 1}$ is the reciprocal of $n_i - 1$, which can be precomputed.
		\end{description}
		\item [The focal individual is heterozygout:] I just have to fill this out and then figure out how to implement it as cleanly as possible.
	\end{description}

\end{description}


\bibliography{anderson}
\bibliographystyle{mychicago}
 \end{document}
